Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-23 07-supplementary-material.qmd
submission materials

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---
```{r}
#| echo: false
#| message: false
#| warning: false

@@

# Supplementary material

```{r echo=FALSE, results='asis'}
#| label: tbl-suple1
#| tbl-cap: "Complete longitudinal multilevel models for market justice preferences"
#| tbl-cap-location: top

ccoef <- list(
  "(Intercept)" = "Intercept",
  "ola2017" = "Wave 2017",
  "ola2018" = "Wave 2018",
  "ola2019" = "Wave 2019",
  "ola2022" = "Wave 2022",
  "ola2023" = "Wave 2023",
  ola_num = "Wave",
  ola_2 = "Wave^2",
  perc_inequality_cwc = "Perception inequality (WE)",
  merit_effort_cwc = "Merit: Effort (WE)",
  merit_talent_cwc = "Merit: Talent (WE)",
  perc_inequality_mean = "Perception inequality (BE)",
  merit_effort_mean = "Merit: Effort (BE)",
  merit_talent_mean = "Merit: Talent (BE)",
  "educUniversitary" = "Universitary education (Ref.= Less than Universitary)",
  quintil1Q2 = "Quintile Q2",
  quintil1Q3 = "Quintile Q3",
  quintil1Q4 = "Quintile Q4",
  quintil1Q5 = "Quintile Q5",
  quintil1QNA = "Quintile no information",
  ess = "Subjective social status",
  ideoCenter = "Center",
  ideoRight = "Right",
  "ideoDoes not identify" = "Does not identify",
  sexFemale = "Female (Ref.= Male)",
  "age30-49" = "Age 30-49",
  "age50-64" = "Age 50-64",
  "age65 or more" = "Age 65 or more"
  )

texreg::knitreg(list(m0, m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c("Model 0",paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6,
                             "Income quintile (Ref.= Quintile 1)" = 16:20,
                             "Political identification (Ref.= Left)" = 21:23,
                             "Age (Ref.= 18-29)" = 25:27),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.7,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"))

>>>>

---
title: "Supplementary material"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=0.5cm
    template-partials: 
      - title.tex
    keep-tex: true
    number-sections: true
editor: source
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
--- 
```{r}
#| echo: false
#| message: false
#| warning: false

@@

```{r echo=FALSE, results='asis'}
ccoef <- list(
  "(Intercept)" = "Intercept",
  "ola2017" = "Wave 2017",
  "ola2018" = "Wave 2018",
  "ola2019" = "Wave 2019",
  "ola2022" = "Wave 2022",
  "ola2023" = "Wave 2023",
  ola_num = "Wave",
  ola_2 = "Wave^2",
  perc_inequality_cwc = "Perception inequality (WE)",
  merit_effort_cwc = "Merit: Effort (WE)",
  merit_talent_cwc = "Merit: Talent (WE)",
  perc_inequality_mean = "Perception inequality (BE)",
  merit_effort_mean = "Merit: Effort (BE)",
  merit_talent_mean = "Merit: Talent (BE)",
  "educUniversitary" = "Universitary education (Ref.= Less than Universitary)",
  quintil1Q2 = "Quintile Q2",
  quintil1Q3 = "Quintile Q3",
  quintil1Q4 = "Quintile Q4",
  quintil1Q5 = "Quintile Q5",
  quintil1QNA = "Quintile no information",
  ess = "Subjective social status",
  ideoCenter = "Center",
  ideoRight = "Right",
  "ideoDoes not identify" = "Does not identify",
  sexFemale = "Female (Ref.= Male)",
  "age30-49" = "Age 30-49",
  "age50-64" = "Age 50-64",
  "age65 or more" = "Age 65 or more"
  )

texreg::texreg(list(m0, m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c("Model 0",paste0("Model ", seq(1:7))),
               caption = "Complete longitudinal multilevel models for market justice preferences",
               caption.above = T,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6,
                             "Income quintile (Ref.= Quintile 1)" = 16:20,
                             "Political identification (Ref.= Left)" = 22:24,
                             "Age (Ref.= 18-29)" = 26:28),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.5,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"))

### 2025-05-23 cover_letter.tex
submission materials

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle

\setstretch{1.15}
\section{Authors Identification}\label{authors-identification}

@@

Julio Iturra, Bremen International Graduate School of Social Sciences,
Bremen, Germany. ORCID: 0000-0003-1993-3479

\section{Exclusive Submission
Statement}\label{exclusive-submission-statement}

@@

\section{Acknowledgments}\label{acknowledgments}

The authors recognize the support of the National Research and
Development Agency (ANID) through the grant from the National Fund for
Scientific and Technological Development (FONDECYT) N° 1210847; from the
Center for Conflict and Social Cohesion Studies (COES) grant Financing
Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and
Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS)
NCS2022\_046.

\section{Authors' Contributions}\label{authors-contributions}

@@

\section{Conflict of Interest}\label{conflict-of-interest}

@@

\section{Funding}\label{funding}

This research received funding from ANID/FONDECYT grant 1210847 and the
Center for Conflict and Social Cohesion Studies (COES) through
ANID/FONDAP grant 1513009.

\section{Data Avaibility Statement}\label{data-avaibility-statement}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

\setstretch{1.15}
\hypertarget{authors-identification}{%
\section{Authors Identification}\label{authors-identification}}

@@

Julio Iturra-Sanhueza, Bremen International Graduate School of Social
Sciences, Bremen, Germany. ORCID: 0000-0003-1993-3479

\hypertarget{exclusive-submission-statement}{%
\section{Exclusive Submission
Statement}\label{exclusive-submission-statement}}

@@

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

The authors recognize the support of the National Research and
Development Agency (ANID) through the grant from the National Fund for
Scientific and Technological Development (FONDECYT) N° 1250518; from the
Center for Conflict and Social Cohesion Studies (COES) grant Financing
Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and
Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS)
NCS2022\_046.

\hypertarget{authors-contributions}{%
\section{Authors' Contributions}\label{authors-contributions}}

@@

\hypertarget{conflict-of-interest}{%
\section{Conflict of Interest}\label{conflict-of-interest}}

@@

\hypertarget{funding}{%
\section{Funding}\label{funding}}

This research received funding from ANID/FONDECYT grant 1250518 and the
Center for Conflict and Social Cohesion Studies (COES) through
ANID/FONDAP grant 1523A0005.

\hypertarget{data-avaibility-statement}{%
\section{Data Avaibility Statement}\label{data-avaibility-statement}}

### 2025-05-23 paper-blinded.qmd
submission materials

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "input/bib/harvard.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra-Sanhueza
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "input/bib/harvard.csl"
# author:
#   - name: Juan Carlos Castillo
#     corresponding: true
#     orcid: 0000-0003-1265-7854
#     email: juancastillov@uchile.cl
#     affiliations:
#       - ref: 1
#       - ref: 2
#       - ref: 3
#   - name: Andreas Laffert
#     orcid: 0000-0002-9008-2454
#     email: alaffertt@estudiante.uc.cl 
#     affiliations:
#       - ref: 4
#   - name: Kevin Carrasco
#     orcid: 0000-0002-9348-8361
#     email: kevin.carrasco@ug.uchile.cl
#     affiliations:
#       - ref: 2
#   - name: Julio Iturra-Sanhueza
#     orcid: 0000-0003-1993-3479
#     email: julioiturrasanhueza@gmail.com 
#     affiliations:
#       - ref: 5     
# affiliations: 
#   - id: 1
#     name: Departamento de Sociología, Universidad de Chile
#   - id: 2
#     name: Centro de estudios del conflicto y cohesión social (COES)
#   - id: 3
#     name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
#   - id: 4
#     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
#   - id: 5
#     name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---
